Smart Drill-Down

### 2015-07-11 TableSummarization/Paper/algorithms.tex
added stuff from tech report

\techreporttext{
\begin{algorithm}
\scriptsize
\KwIn{$k$ (Number of rules required), $T$ (database table), $m_w$ (max weight), $W$ (weight function)}
\KwOut{$S$ (Solution set of rules)}
$S = \phi$

@@

$S = S \cup \left\lbrace R_m \right\rbrace$
}
\Return $S$
\caption{Greedy Algorithm for Problem~\ref{prob:optimal-rule-set}\label{algo:best-rule-set}}
\end{algorithm}
}

@@

\techreporttext{
\begin{algorithm}
\scriptsize
\KwIn{$S$ (Current solution set), $T$ (database table), $m_w$ (max weight), $W$ (weight function)}
\KwOut{$R_m$ (Rule which adds the highest marginal value among rules with weight $\leq m_w$)}
$H = 0$ \tcc*{Threshold for deciding if to count for a rule.}
$C = C_o = C_n = \phi$ \tcc*{Set of all, old and new candidate rules respectively. }
\For{$j$ from $1$ to number of columns in $T$}{
\If {$j = 1$}{
$C_n = \text{ all rules of size } 1$
}
\Else{ 
$C_n = \text{all size-}i$ super-rules of rules from $C_o$
}
\ForEach {$R \in C_n$}{
$M =\infty$ \tcc*{Upper bound on marginal value count of $R$}
\ForEach {$R$-sub-rule $R^{\prime} \in C$}{  
$M = \text{min}(M, \text{MarginalVal}(R^{\prime}) + \text{Count}(R^{\prime})(m_w - W(R^{\prime}))$
}
\If {$(M < H)$}{ 
$C_n = C_n \setminus \left\lbrace R \right\rbrace$ \tcc{Delete $R$ if its max count is too small for $R$ to be in the solution}
}
}
\If {$C_n = \phi$}{
break;
}
\ForEach {$R \in C_n$}{
Count$(R) = 0$ \tcc*{Initialize}
MarginalValue$(R) = 0$ \tcc*{Initialize}
}
\ForEach {$t \in T$}{
Let $R_S$ be the highest weight rule in $S$ that covers $t$

@@

$H = \textrm{max}_{R \in C}(\text{MarginalValue}(R))$
}
\Return $\textrm{argmax}_{r \in C} \text{MarginalValue}(r)$
\caption{Find best marginal rule\label{algo:best-marginal-rule}}
\end{algorithm}
}

@@

\stitle{Parametrized Algorithm:} \techreporttext{Our algorithm pseudo-code is given in the box labelled Algorithm~\ref{algo:best-rule-set}.} \papertext{We describe our algorithm in detail below. The pseudo-code for our algorithm is in the technical report~\cite{tr}.} We call our algorithm {\em BRS} (for \textbf{B}est \textbf{R}ule \textbf{S}et). BRS takes four parameters as input: the table $T$, the number $k$ of rules required in the final solution list, a parameter $m_w$ (which we describe in the next paragraph), and the weight function $W$.

@@

We now describe the function to find the best marginal rule. \techreporttext{The pseudo-code for the function is in the box titled Algorithm~\ref{algo:best-marginal-rule}.} \papertext{The pseudo-code for the function is in the technical report.} The function maintains a threshold $H$, which is the highest marginal value that has been found for any rule so far. The function makes several iterations \techreporttext{(Step $3$)}, counting marginal values for size $j$ rules in the $j^{th}$ iteration. We maintain three sets of rules. $C$ is the set of all rules whose marginal values have been counted in all previous iterations. $C_n$ is the set of rules whose marginal values are going to be counted in the current pass. And $C_o$ is the set of rules whose marginal values were counted in the previous iteration. For the first pass, we set $C_n$ to be all rules of size $1$. Then we compute marginal values for those rules, and set $C = C_o = C_n$.

For the second pass onwards, we are more selective about which rules to consider for marginal value evaluation. We first set $C_n$ to be the set of rules of size $j$ which are super-rules of rules from $C_o$. Then for each rule $r$ from $C_n$, we consider the known marginal values of its sub-rules from $C$, and use them to upper-bound the marginal value of all super-rules of $r$\techreporttext{, as shown in Step 3.3.2}. Then we delete from $C_n$ the rules whose marginal value upper bound is less than the currently known best marginal value, since they have no chance of being returned as the best marginal rule. Then we make as actual pass through the table to compute the marginal value of the rules in $C_n$\techreporttext{, as shown in Step 3.5}. If in any round, the $C_n$ obtained after deleting rules is empty, then we terminate the algorithm and return the highest value rule.

>>>>

\begin{algorithm}
\scriptsize
\KwIn{$k$ (Number of rules required), $T$ (database table), $m_w$ (max weight), $W$ (weight function)}
\KwOut{$S$ (Solution set of rules)}
$S = \phi$

@@

$S = S \cup \left\lbrace R_m \right\rbrace$
}
\Return $S$
\caption{Greedy Algorithm for Problem~\ref{prob:optimal-rule-set}\label{algo:best-rule-set}}
\end{algorithm}

@@

\begin{algorithm}
\scriptsize
\KwIn{$S$ (Current solution set), $T$ (database table), $m_w$ (max weight), $W$ (weight function)}
\KwOut{$R_m$ (Rule which adds the highest marginal value among rules with weight $\leq m_w$)}
$H = 0$ \tcc*{Threshold for deciding if to count for a rule.}
$C = C_o = C_n = \phi$ \tcc*{Set of all, old and new candidate rules respectively. }
\For{$j$ from $1$ to number of columns in $T$}{
\If {$j = 1$}{
$C_n = \text{ all rules of size } 1$
}
\Else{ 
$C_n = \text{all size-}i$ super-rules of rules from $C_o$
}
\ForEach {$R \in C_n$}{
$M =\infty$ \tcc*{Upper bound on marginal value count of $R$}
\ForEach {$R$-sub-rule $R^{\prime} \in C$}{  
$M = \text{min}(M, \text{MarginalVal}(R^{\prime}) + \text{Count}(R^{\prime})(m_w - W(R^{\prime}))$
}
\If {$(M < H)$}{ 
$C_n = C_n \setminus \left\lbrace R \right\rbrace$ \tcc{Delete $R$ if its max count is too small for $R$ to be in the solution}
}
}
\If {$C_n = \phi$}{
break;
}
\ForEach {$R \in C_n$}{
Count$(R) = 0$ \tcc*{Initialize}
MarginalValue$(R) = 0$ \tcc*{Initialize}
}
\ForEach {$t \in T$}{
Let $R_S$ be the highest weight rule in $S$ that covers $t$

@@

$H = \textrm{max}_{R \in C}(\text{MarginalValue}(R))$
}
\Return $\textrm{argmax}_{r \in C} \text{MarginalValue}(r)$
\caption{Find best marginal rule\label{algo:best-marginal-rule}}
\end{algorithm}

@@

\stitle{Parametrized Algorithm:} Our algorithm pseudo-code is given in the box labelled Algorithm~\ref{algo:best-rule-set}. We call our algorithm {\em BRS} (for \textbf{B}est \textbf{R}ule \textbf{S}et). BRS takes four parameters as input: the table $T$, the number $k$ of rules required in the final solution list, a parameter $m_w$ (which we describe in the next paragraph), and the weight function $W$.

@@

We now describe the function to find the best marginal rule. The pseudo-code for the function is in the box titled Algorithm~\ref{algo:best-marginal-rule}. The function maintains a threshold $H$, which is the highest marginal value that has been found for any rule so far. The function makes several iterations (Step $3$), counting marginal values for size $j$ rules in the $j^{th}$ iteration. We maintain three sets of rules. $C$ is the set of all rules whose marginal values have been counted in all previous iterations. $C_n$ is the set of rules whose marginal values are going to be counted in the current pass. And $C_o$ is the set of rules whose marginal values were counted in the previous iteration. For the first pass, we set $C_n$ to be all rules of size $1$. Then we compute marginal values for those rules, and set $C = C_o = C_n$.

For the second pass onwards, we are more selective about which rules to consider for marginal value evaluation. We first set $C_n$ to be the set of rules of size $j$ which are super-rules of rules from $C_o$. Then for each rule $r$ from $C_n$, we consider the known marginal values of its sub-rules from $C$, and use them to upper-bound the marginal value of all super-rules of $r$, as shown in Step 3.3.2. Then we delete from $C_n$ the rules whose marginal value upper bound is less than the currently known best marginal value, since they have no chance of being returned as the best marginal rule. Then we make as actual pass through the table to compute the marginal value of the rules in $C_n$, as shown in Step 3.5. If in any round, the $C_n$ obtained after deleting rules is empty, then we terminate the algorithm and return the highest value rule.

### 2015-07-11 TableSummarization/Paper/related.tex
changes to related work and intro
added stuff from tech report

% All the existing works solve a fixed optimization problem, whereas  we focus on finding an optimal summary for a flexible user-specified weighting function.

>>>>

There has also been plenty of work on pattern mining. Several works~\cite{Vreeken:2011:KMI:1969593.1969615,Bringmann:2007:ICDM,Yan:2005:SIP:1081870.1081907} propose non-interactive schemes that attempt to find a one shot summary of the table. These schemes usually consume a large amount of time processing the whole table, rather than allowing the user to slowly steer into portions of interest. In contrast, our work is interactive, and includes a smart memory manager that can use limited memory effectively while preparing for future requests.

Our Smart Drill-Down operator is tunable because of the flexible weighting function, but the monotonicity of the weighting function and the use of $MCount$, still make it possible for us to get an approximate optimality guarantee for the rules we display. In contrast, much of the existing pattern mining work~\cite{Goethals:2011:MFI:2020408.2020529,Tatti:2014:FRI:2676651.2656261,DeBie:2010:FMI:1816112.1816117} is not not tunable enough, providing only a fixed set of interestingess parameters. On the other hand, reference~\cite{Leeuwen:2012:DSS:2347179.2347243} allows a fully general scoring function, necessitating the use of heuristics with no optimality guarantees, and very time consuming algorithms. 
A lot of pattern mining work~\cite{Goethals:2011:MFI:2020408.2020529,Yan:2005:SIP:1081870.1081907,Vreeken:2011:KMI:1969593.1969615} also focuses on itemsets rather than Relational Data, which does not allow the user to express interest in certain `columns' over others.

### 2015-07-11 TableSummarization/Paper/sampling.tex
small changes throughout
added stuff from tech report

\papertext{
BRS makes multiple passes over the table in order to determine the best set of rules to display. This can be slow when the table is too large to fit in main memory. We can reduce the response time of smart drill down by running BRS on a sample of the table instead, trading off accuracy of our rules for performance. If we had obtained a sample $s$ by selecting each table tuple with probability $p$, and run BRS on $s$, then we multiply the count of each rule found by BRS, by $\frac{1}{p}$ to estimate its count over the full table.

In Section~\ref{sec:sampling_algorithms}, we describe the problem of optimally allocating memory to different samples. We show that the problem is NP-Hard, and describe an approximate solution. In Section~\ref{sec:sample-handler}, we describe a component of the system called the {\em SampleHandler}, which is responsible for creating and maintaining multiple samples of different parts of the table in memory, and creating temporary samples for BRS to process. Detailed descriptions of the sampling problem and the SampleHandler are in the technical report. The technical report also contains some additional optimizations, and ways to set the minimum sample size.

\subsubsection{Deciding what to sample}\label{sec:sampling_algorithms}
We are given a memory capacity $M$, and a minimum sample size $minSS$, both specified by the user. $minSS$ is the minimum number of tuples on which we are allowed to run BRS, without accessing the hard disk. A higher value of $minSS$ increases both accuracy and computation cost of our system.

At any point, we have a tree $U$ of rules displayed to the user. Initially, the tree consists of a single node corresponding to the empty rule. When the user drills down on a rule $r$, the sub-rules of $r$ obtained by running BRS are added as children of node $r$. 
Our system maintains multiple samples in memory, with one sample per rule in $U$. Specifically, for each rule $r \in U$, we choose an integer $n_r$, and create a sample $s_r$ consisting of $n_r$ uniformly randomly chosen tuples covered by $r$ from the table. Because of the memory constraint, we must have $\sum_{r\in U} n_r \leq M$.

When a user attempts to drill down on $r$, the SampleHandler takes all $n_r$ tuples from $s_r$, and also tuples covered by $r$ from samples $s_{r'}$ for all $r' \in U$ that are sub-rules of $r$. If the total number of such tuples is $\geq minSS$, then we run BRS on that set of tuples to perform the drill down. Note that this set of tuples forms a uniformly random sample of tuples covered by $r$. If not, then we need to access the hard disk to obtain more tuples covered by $r$.

Leaves of tree $U$ correspond to rules that may be drilled down on next. We assume there is a probability distribution over leaves, which assigns a probability that each leaf may be drilled down on next. This can be a uniform distribution, or a machine learned distribution using past user data. We aim to set sample sizes $n_r$ so as to maximize the probability that the next drill down can be performed without accessing the hard disk.

If $r'$ is a sub-rule of $r$, and covers $x$ times as many rules as $r$, then it means that when drilling down on $r$, $s_{r'}$ can contribute around $\frac{n_{r'}}{x}$ tuples to the sample for $r$. We denote the ratio of selectivities $x$ by $S(r',r)$. $S(r',r)$ is defined to be $0$ if $r'$ is not a sub-rule of $r$. If $r$ is to be drilled down on next, the total number of sample tuples we will have for $r$ from all existing samples is given by $ess(r) = \sum_{r'\in U} S(r', r)n_{r'}$. We can drill down on $r$ without accessing hard disk, if $ess(r) \geq minSS$. We now formally define our problem: 
\begin{problem}\label{prob:sample-sizes}
Given a tree of rules $U$ with leaves $L$, a probability distribution $p$ over $L$, an integer $M$, and selectivity ratio $S(r_1, r_2)$ for each $r_1, r_2 \in U$, 
choose an integer $n_r \geq 0$ for each $r \in U$ so as to $\textrm{maximize}$ :
$$\sum_{r^{\prime} \in L} p_{r^{\prime}}I_{[ess(r^{\prime}) \geq minSS]}$$
where the $I$'s are indicator variables, subject to :
$\sum_{r \in U} n_r \leq M$
\end{problem}
Problem~\ref{prob:sample-sizes} is non-linear and non-convex because of the indicator variables. 
%If the tree $U$ is not too big, then it might be feasible to use an exponential algorithm to solve the problem. 
We can show that Problem~\ref{prob:sample-sizes} is {\sc NP-Hard} using a reduction from the knapsack problem. 
\begin{lemma}\label{lemma:sampling-hardness}
Problem~\ref{prob:sample-sizes} is {\sc NP-Hard}.
\end{lemma}
\begin{proof}(Sketch)
Suppose we are given an instance of the knapsack problem with $m$ objects, with the $i^{th}$ object having weight $w_i$ and value $v_i$. We are also given a weight limit $W$, and our objective is to choose a set of objects that maximizes value and has total weight $< W$. We will reduce this instance to an instance of Problem~\ref{prob:sample-sizes}.

We first scale the $w_i$s and $W$ such that all $w_i$s are $< 1$. For Problem~\ref{prob:sample-sizes}, we set $M$ to $(m + W) \times minSS$. Tree $U$ has $m$ special nodes $r_1, r_2, ... r_m$, and each $r_i$ has two leaf children $r_{i,1}, r_{i,2}$. All other leaves have expansion probability $0$. 
The $S$ values are such that $\forall i \in \left\lbrace 1, 2, ..m\right\rbrace, j \in \left\lbrace 1, 2 \right\rbrace : (S(x, r_{i,j}) \neq 0 \Rightarrow x = r_{i,j} || x = r_i)$. In reality, the $S$ values cannot be exactly zero, but can be made small enough for all practical purposes. Thus, $\forall 1 \leq i \leq m, j \in \left\lbrace 1,2 \right\rbrace : ess(r_{i,j}) = n_{r_{i,j}} + n_{r_i}S(r_i, r_{i,j})$. In addition, $S(r_i, r_{i,1}) = 1$, and $S(r_i, r_{i,2}) = 1 - w_i$. Finally, $\forall i : p_{r_{i,1}} = \frac{2}{2m+1}$ and $p_{r_{i,2}} = \frac{v_i}{(2m+1)\sum_{j=1}^{m}v_i}$. So in any optimal solution, $\forall i : ess(r_{i,1}) = minSS$, and we've to decide which $i$'s also have $ess(r_{i,2}) = minSS$. Having $ess(r_{i,2}) = minSS$ requires consuming $w_i \times minSS$ extra memory and gives an extra $\frac{v_i}{(2m+1)(\sum_{j=1}^{m}v_j)}$ probability value. Thus, having $ess(r_{i,2}) = minSS$ is equivalent to picking object $i$ from the knapsack problem. Solving Problem~\ref{prob:sample-sizes} with the above $U$, $S$, $p$ and picking the set of $i$'s for which $ess(r_{i,2}) = minSS$ gives a solution to the instance of the knapsack problem. 
\end{proof}
A more detailed proof sketch is given in the technical report.
}

\techreporttext{
Our greedy algorithm needs to make multiple passes over the entire table in order to find counts of rules. These passes can be expensive if the table is large and the table does not fit in main memory. If we want exact counts for rules, we have no choice but to read the entire table.

@@

We are given a memory capacity or budget $M$, and a minimum sample size $minSS$ (both specified by the user).
The parameter $minSS$ is the minimum number of sample tuples needed
such that we can use the sample in memory 
instead of having to resort to the entire table stored
on hard disk. 
This parameter determines how accurate our count estimates will be, and also
how quickly we can return results to the user. 
\papertext{We describe how to set $minSS$ in the technical report.}

@@

We first scale the $w_i$s and $W$ such that all $w_i$s are $< 1$, without effectively changing the problem. For Problem~\ref{prob:sample-sizes}, we set $M$ to $(m + W) \times minSS$. Tree $U$ has $m$ special nodes $r_1, r_2, ... r_m$, and each $r_i$ has two children $r_{i,1}, r_{i,2}$. These $2m$ children are all leaves, and all leaves other than these have expansion probability $0$. 
The $S$ values are such that $\forall i \in \left\lbrace 1, 2, ..m\right\rbrace, j \in \left\lbrace 1, 2 \right\rbrace : (S(x, r_{i,j}) \neq 0 \Rightarrow x = r_{i,j} || x = r_i)$. In reality, the $S$ values cannot be exactly zero when the first argument is the trivial rule, but we can make it small enough such that any optimal solution will set $n_r = 0$ when $r$ is not a special node or its child. Therefore, each leaf $r_{i,j}$ gets tuples either from its own sample, or the sample of its parent, and from nowhere else. Thus, $ess(r_{i,j}) = n_{r_{i,j}} + n_{r_i}S(r_i, r_{i,j}) \forall 1 \leq i \leq m, j \in \left\lbrace 1,2 \right\rbrace$. In addition, $S(r_i, r_{i,1}) = 1$ (again, it cannot be exactly $1$, but can be brought arbitrarily close to $1$), and $S(r_i, r_{i,2}) = 1 - w_i$. Finally, for each i, $p_{r_{i,1}} = \frac{2}{2m+1}$ and $p_{r_{i,2}} = \frac{v_i}{(2m+1)\sum_{j=1}^{m}v_i}$. Thus, each individual $p_{r_{i,1}}$ value is higher than all $p_{r_{i,2}}$ values combined, and $M$ is high enough to cover all $r_{i,1}$. As a result, in any optimal solution, $ess(r_{i,1})$ must be equal to $minSS$ for all $i$, and we're left to decide which $i$'s should also have $ess(r_{i,2}) = minSS$. For all $i$, we must have either $n_{r_i} = n_{r_{i,2}} = 0 \land n_{r_{i,1}} = minSS$ (iff $ess(r_{i,2}) < minSS$) or $n_{r_i} = minSS \land n_{r_{i,1}} = 0 \land n_{r_{i,2}} = minSS (1 - S(r_i, r_{i,2})) = minSS \times w_i$ (iff $ess(r_{i,2}) = minSS$). The latter option consumes $w_i \times minSS$ extra memory and gives an extra $\frac{v_i}{(2m+1)(\sum_{j=1}^{m}v_j)}$ probability value. Thus, having $ess(r_{i,2}) = minSS$ is equivalent to picking object $i$ from the knapsack problem (as it consumes additional memory $\propto w_i$ and gives additional probability value $\propto v_i$). Moreover, the additional memory available (on top of the $m \times minSS$ required to cover all $r_{i,1}$s) is $W\times minSS$. Hence, solving Problem~\ref{prob:sample-sizes} with the above $U$, $S$, $p$ and picking the set of $i$'s for which $ess(r_{i,2}) = minSS$ gives a solution to the instance of the knapsack problem. 
\end{proof}
}

@@

\papertext{We present the idea behind the approximate solution below. A more detailed description can be found in the technical report. The solution consists of two main steps:
\squishlist
\item \textbf{Locally Optimal Solutions :} For each rule $r_0 \in U \setminus L$, let $M_{r_0}$ be the set consisting of $r_0$ and its leaf children. We consider the number of tuples assigned to members of $M_{r_0}$. Each such assignment of numbers has a `cost' equal to total number of tuples assigned, and `value' equal to sum of probabilities of $M_{r_0}$ members with $ess > minSS$. We consider all `locally optimal' assignments, where probability value cannot be increased without increasing the cost.
\item \textbf{Combining Solutions :} Then, we use dynamic programming to pick one optimal solution per $r_0$ so as to get maximum probability value with the given cost constraint. We discretize costs, say to have granularity $100$. Let $M_{0}, M_{1}, ... M_{D}$ be the sets for which we have found locally optimal solutions. We make $D$ iterations in all, where in the $i^{th}$ iteration, we try different locally optimal solutions for $M_i$ and use it to update the maximum attainable value for each cost. 
\squishend
}

\techreporttext{Now consider a rule in $r_0 \in U \setminus L$ along with all its children. Let $M_{r_0}$ denote the set containing $r_0$ and all its leaf children. Then, by our simplification, the number of tuples $n_{r^{\prime}}$ for any rule in $r^{\prime} \in M_{r_0}$ only affects the $ess$ value of rules in $M_{r_0}$. This allows us to effectively split the problem into multiple subproblems,  with one subproblem per $M_{r_0}$. Thus, for each non-leaf rule $r_0$ and all its children, we compute all `locally optimal' assignments of $n_r \mid r \in M_{r_0}$. Locally optimal means that we cannot get a higher value of `probability value' $\sum_{r \in M_{r_0}} p_rI_{ess(r) \geq minSS}$ for the same `sampling cost' $\sum_{r\in M_{r_0}} n_r$. Then, we can use dynamic programming to combine the locally optimal solutions of different $M_{r_0}$s. We describe both these steps in detail below:

@@

Then next step is to combine the solutions using dynamic programming. Let the $M$ sets be called $M_0, M_1, ... M_D$. Let our possible sample sizes range from $0$ to $\mathcal{S}$. The number of sample sizes can be pretty large ($\mathcal{S}$), but we can make it smaller by discretizing the sample sizes, say to have granularity $100$. Then we create a $D \times \mathcal{S}$ array $A$. The value $A\left[i\right]\left[j\right]$ contains the best probability score we can get from $M_0, M_1, ... M_i$ with total sample size at most $j$. We can populate $A\left[0\right]\left[j\right] \forall j$ using the locally optimal solutions for $M_0$. Let $E_{i+1}$ denote the set of locally optimal solutions for $M_{i+1}$. Then we have,
$$A\left[i+1\right] \left[j \right] = \textrm{max} (A\left[i\right]\left[j\right], \textrm{max}_{e \in E_{i+1}}(A\left[i\right]\left[j-S(e)\right] + P(e)))$$
This can be solved using dynamic programming, in $O(D\mathcal{S}3^d)$ time. 
}

@@

\subsubsection{Design of the SampleHandler}\label{sec:sample-handler}
\papertext{
We now briefly describe the design of the {\em SampleHandler}, which given
a certain memory capacity $M$, and a minimum sample size $minSS$,
creates, maintains, retrieves, and removes samples, all in
response to user interactions on the table. A detailed description, with some potential optimizations, can be found in the technical report.

@@

\item \textbf{Create:} If \textbf{Combine} doesn't work either, then the SampleHandler needs to create a new sample $s$ with $f_s = r$ by making a pass through the table.  Since accessing the hard disk is expensive, it also creates samples for rules other than $r$ and augments existing samples in the same pass. It first deletes all existing samples, then uses the algorithm from Section~\ref{sec:sampling_algorithms} to determine how many tuples to sample for each rule, and uses reservoir sampling~\cite{maibdr1983,Vitter:1985:RSR:3147.3165} to create a uniformly random sample of a given size for each rule.
\squishend
}

\techreporttext{
We now describe the design of the {\em SampleHandler}, which given
a certain memory capacity $M$, and a minimum sample size $minSS$,
creates, maintains, retrieves, and removes samples, all in
response to user interactions on the table.
It uses the algorithms from Section~\ref{sec:sampling_algorithms} 
to decide which samples to create, as we will see below.

@@

\smallskip
\noindent \textbf{Pre-fetching:} When the user clicks on rule $r$ (on the rule itself or on a $\star$ in the rule), we need to get a sample, run the BRS, and display a rule-list to the user. If we use \textbf{Find} or \textbf{Combine}, then we can display the rule-list much faster because we don't have to read the entire table. But after expanding $r$, there is a high chance that the user goes further and drills down on one of the sub-rules $r^{\prime}$ of $r$. We may not be able to use \textbf{Find} or \textbf{Combine} on $r^{\prime}$ with the existing samples. So while the user is busy reading the current rule-list obtained from drilling down on $r$, we can start running the algorithm from Section~\ref{sec:sampling_algorithms} in the background, and then making a pass through the table to create a new samples. That way, when the user expands the next rule $r^{\prime}$, there will be a high chance of a sample being pre-fetched for $r^{\prime}$, increasing the chance that we can use \textbf{Find} or \textbf{Combine} on $r^{\prime}$ and reducing our response time.
In addition, while we are making the pass in the background, we can find the exact counts for currently displayed rules (which only have estimated counts shown), and update them when our pass is complete.
}

>>>>

Our greedy algorithm needs to make multiple passes over the entire table in order to find counts of rules. These passes can be expensive if the table is large and the table does not fit in main memory. If we want exact counts for rules, we have no choice but to read the entire table.

@@

We are given a memory capacity or budget $M$, and a minimum sample size $minSS$ (both specified by the user).
The parameter $minSS$ is the minimum number of sample tuples needed
such that we can use the sample in memory 
instead of having to resort to the entire table stored
on hard disk. 
This parameter determines how accurate our count estimates will be, and also
how quickly we can return results to the user.

@@

We first scale the $w_i$s and $W$ such that all $w_i$s are $< 1$, without effectively changing the problem. For Problem~\ref{prob:sample-sizes}, we set $M$ to $(m + W) \times minSS$. Tree $U$ has $m$ special nodes $r_1, r_2, ... r_m$, and each $r_i$ has two children $r_{i,1}, r_{i,2}$. These $2m$ children are all leaves, and all leaves other than these have expansion probability $0$. 
The $S$ values are such that $\forall i \in \left\lbrace 1, 2, ..m\right\rbrace, j \in \left\lbrace 1, 2 \right\rbrace : (S(x, r_{i,j}) \neq 0 \Rightarrow x = r_{i,j} || x = r_i)$. In reality, the $S$ values cannot be exactly zero when the first argument is the trivial rule, but we can make it small enough such that any optimal solution will set $n_r = 0$ when $r$ is not a special node or its child. Therefore, each leaf $r_{i,j}$ gets tuples either from its own sample, or the sample of its parent, and from nowhere else. Thus, $ess(r_{i,j}) = n_{r_{i,j}} + n_{r_i}S(r_i, r_{i,j}) \forall 1 \leq i \leq m, j \in \left\lbrace 1,2 \right\rbrace$. In addition, $S(r_i, r_{i,1}) = 1$ (again, it cannot be exactly $1$, but can be brought arbitrarily close to $1$), and $S(r_i, r_{i,2}) = 1 - w_i$. Finally, for each i, $p_{r_{i,1}} = \frac{2}{2m+1}$ and $p_{r_{i,2}} = \frac{v_i}{(2m+1)\sum_{j=1}^{m}v_i}$. Thus, each individual $p_{r_{i,1}}$ value is higher than all $p_{r_{i,2}}$ values combined, and $M$ is high enough to cover all $r_{i,1}$. As a result, in any optimal solution, $ess(r_{i,1})$ must be equal to $minSS$ for all $i$, and we're left to decide which $i$'s should also have $ess(r_{i,2}) = minSS$. For all $i$, we must have either $n_{r_i} = n_{r_{i,2}} = 0 \land n_{r_{i,1}} = minSS$ (iff $ess(r_{i,2}) < minSS$) or $n_{r_i} = minSS \land n_{r_{i,1}} = 0 \land n_{r_{i,2}} = minSS (1 - S(r_i, r_{i,2})) = minSS \times w_i$ (iff $ess(r_{i,2}) = minSS$). The latter option consumes $w_i \times minSS$ extra memory and gives an extra $\frac{v_i}{(2m+1)(\sum_{j=1}^{m}v_j)}$ probability value. Thus, having $ess(r_{i,2}) = minSS$ is equivalent to picking object $i$ from the knapsack problem (as it consumes additional memory $\propto w_i$ and gives additional probability value $\propto v_i$). Moreover, the additional memory available (on top of the $m \times minSS$ required to cover all $r_{i,1}$s) is $W\times minSS$. Hence, solving Problem~\ref{prob:sample-sizes} with the above $U$, $S$, $p$ and picking the set of $i$'s for which $ess(r_{i,2}) = minSS$ gives a solution to the instance of the knapsack problem. 
\end{proof}

@@

\begin{comment}
\papertext{We present the idea behind the approximate solution below. A more detailed description can be found in the technical report. The solution consists of two main steps:
\squishlist
\item \textbf{Locally Optimal Solutions :} For each rule $r_0 \in U \setminus L$, let $M_{r_0}$ be the set consisting of $r_0$ and its leaf children. We consider the number of tuples assigned to members of $M_{r_0}$. Each such assignment of numbers has a `cost' equal to total number of tuples assigned, and `value' equal to sum of probabilities of $M_{r_0}$ members with $ess > minSS$. We consider all `locally optimal' assignments, where probability value cannot be increased without increasing the cost.
\item \textbf{Combining Solutions :} Then, we use dynamic programming to pick one optimal solution per $r_0$ so as to get maximum probability value with the given cost constraint. We discretize costs, say to have granularity $100$. Let $M_{0}, M_{1}, ... M_{D}$ be the sets for which we have found locally optimal solutions. We make $D$ iterations in all, where in the $i^{th}$ iteration, we try different locally optimal solutions for $M_i$ and use it to update the maximum attainable value for each cost. 
\squishend
}
\end{comment}

Now consider a rule in $r_0 \in U \setminus L$ along with all its children. Let $M_{r_0}$ denote the set containing $r_0$ and all its leaf children. Then, by our simplification, the number of tuples $n_{r^{\prime}}$ for any rule in $r^{\prime} \in M_{r_0}$ only affects the $ess$ value of rules in $M_{r_0}$. This allows us to effectively split the problem into multiple subproblems,  with one subproblem per $M_{r_0}$. Thus, for each non-leaf rule $r_0$ and all its children, we compute all `locally optimal' assignments of $n_r \mid r \in M_{r_0}$. Locally optimal means that we cannot get a higher value of `probability value' $\sum_{r \in M_{r_0}} p_rI_{ess(r) \geq minSS}$ for the same `sampling cost' $\sum_{r\in M_{r_0}} n_r$. Then, we can use dynamic programming to combine the locally optimal solutions of different $M_{r_0}$s. We describe both these steps in detail below:

@@

Then next step is to combine the solutions using dynamic programming. Let the $M$ sets be called $M_0, M_1, ... M_D$. Let our possible sample sizes range from $0$ to $\mathcal{S}$. The number of sample sizes can be pretty large ($\mathcal{S}$), but we can make it smaller by discretizing the sample sizes, say to have granularity $100$. Then we create a $D \times \mathcal{S}$ array $A$. The value $A\left[i\right]\left[j\right]$ contains the best probability score we can get from $M_0, M_1, ... M_i$ with total sample size at most $j$. We can populate $A\left[0\right]\left[j\right] \forall j$ using the locally optimal solutions for $M_0$. Let $E_{i+1}$ denote the set of locally optimal solutions for $M_{i+1}$. Then we have,
$$A\left[i+1\right] \left[j \right] = \textrm{max} (A\left[i\right]\left[j\right], \textrm{max}_{e \in E_{i+1}}(A\left[i\right]\left[j-S(e)\right] + P(e)))$$
This can be solved using dynamic programming, in $O(D\mathcal{S}3^d)$ time.

@@

\subsubsection{Design of the SampleHandler}\label{sec:sample-handler}
\begin{comment}
\papertext{
We now briefly describe the design of the {\em SampleHandler}, which given
a certain memory capacity $M$, and a minimum sample size $minSS$,
creates, maintains, retrieves, and removes samples, all in
response to user interactions on the table. A detailed description, with some potential optimizations, can be found in the technical report.

@@

\item \textbf{Create:} If \textbf{Combine} doesn't work either, then the SampleHandler needs to create a new sample $s$ with $f_s = r$ by making a pass through the table.  Since accessing the hard disk is expensive, it also creates samples for rules other than $r$ and augments existing samples in the same pass. It first deletes all existing samples, then uses the algorithm from Section~\ref{sec:sampling_algorithms} to determine how many tuples to sample for each rule, and uses reservoir sampling~\cite{maibdr1983,Vitter:1985:RSR:3147.3165} to create a uniformly random sample of a given size for each rule.
\squishend
}
\end{comment}

%\techreporttext{
We now describe the design of the {\em SampleHandler}, which given
a certain memory capacity $M$, and a minimum sample size $minSS$,
creates, maintains, retrieves, and removes samples, all in
response to user interactions on the table.
It uses the algorithms from Section~\ref{sec:sampling_algorithms} 
to decide which samples to create, as we will see below.

@@

\techreporttext{
\smallskip
\noindent \textbf{Pre-fetching:} When the user clicks on rule $r$ (on the rule itself or on a $\star$ in the rule), we need to get a sample, run the BRS, and display a rule-list to the user. If we use \textbf{Find} or \textbf{Combine}, then we can display the rule-list much faster because we don't have to read the entire table. But after expanding $r$, there is a high chance that the user goes further and drills down on one of the sub-rules $r^{\prime}$ of $r$. We may not be able to use \textbf{Find} or \textbf{Combine} on $r^{\prime}$ with the existing samples. So while the user is busy reading the current rule-list obtained from drilling down on $r$, we can start running the algorithm from Section~\ref{sec:sampling_algorithms} in the background, and then making a pass through the table to create a new samples. That way, when the user expands the next rule $r^{\prime}$, there will be a high chance of a sample being pre-fetched for $r^{\prime}$, increasing the chance that we can use \textbf{Find} or \textbf{Combine} on $r^{\prime}$ and reducing our response time.
In addition, while we are making the pass in the background, we can find the exact counts for currently displayed rules (which only have estimated counts shown), and update them when our pass is complete.
}
